Synthesizing Quotient Lenses

### 2018-03-07 papers/QuotientLenses/main.tex
Polishing in 3.1 and 3.2 -- release locks

\bcp{We need two subsections here --- one on plain old bijective lenses, and
one on bijective lens synthesis.}

@@

The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c_1}{c_2}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
& | \; \normalize{R_1}{R_2}{f} \sep c_1 \semicolon c_2 \sep c_1 \cdot c_2 \sep (c_1 \sep
c_2) \sep c^*,
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.
$\mathcal{L}(R)$ is
the language accepted by the regular expression $R$.

@@

\subsection{Intuitions}
In the \bibtex{} to EndNote example, we encountered the $\project{R}{s}$,

\noindent $\squash{c}{c'}{f}$ and $\perm{c_1, \ldots, c_n}{c}$ QREs, as well as
the regular operators applied to these, as in\bcp{linebreak}

@@

One combinator that we haven't yet introduced\bcp{haven't we? we've
  mentioned it, at least.} is sequential composition.
The $c \semicolon c'$ QRE successively applies
two equivalence relations to a regular language. For example, suppose that
we wish to express a name as $\kw{last\_name \cdot ``,\text{''} \cdot
\text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot last\_name}$ as the
canonical representation. We also wish to project away the extra whitespace in
between \kw{first\_name} and \kw{last\_name} once this identification has been
made. Then we can express this equivalence relation by a composition of a
\kw{squash} QRE and a QRE that projects away the extra whitespace in between
\kw{first\_name} and \kw{last\_name}:

@@

The final combinator is $\normalize{R}{R'}{f}$, which expresses
the equivalence relation that sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QRE
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are either
difficult to define, or that cannot be defined using the other combinators.
\saz{Add something about being ``complete'' here?}\sam{I'm not sure how to do
this here. The precise statement about how normalize is strong enough to
express all QREs requires that we introduce W(c), K(c) and Canonizer(c) which
we haven't introduced at this point in the paper}

Now in order for the programmer to ensure that the {\em normalize} QRE is well
formed, she must define a canonizing function $f$ between two regular
languages and prove that $f$ is surjective and idempotent: the compiler cannot
verify this since this problem is undecidable in general. Consequently, we do
not use the $\normalize{R}{R'}{f}$ QRE in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers. On the other hand, QREs enable us to
express many equivalences that occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

The net effect of these design decisions means that each QRE $c$ enables
us to simultaneously express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Observe that since $\canonizer(c)$ computes $\canonizer(c)(w)$ as the unique
$k$ in $\mathcal{L}(K(c))$ such that $w$ is equivalent to $k$ mod $\eqrel{c}$,
then the equivalence classes of $\eqrel{c}$ are the same as the fibres of
$\canonizer(c)$.

>>>>

\bcp{We need to write two new subsections here --- one on plain old
  bijective lenses, and one on bijective lens synthesis.}

@@

The language of Quotient Regular Expressions (QREs) is given by the grammar
\begin{align*}
c := \; &R \sep \project{R}{s} \sep \squash{c_1}{c_2}{f} \sep
\perm{c_1, \ldots, c_n}{c} \\
& | \; \normalize{R_1}{R_2}{f} \sep c_1 \semicolon c_2 \sep c_1 \cdot c_2 \sep (c_1 \sep
c_2) \sep c^*
\end{align*}
where $c$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.
$\mathcal{L}(R)$ is
the language accepted by the regular expression $R$.

@@

\subsection{Intuitions}
In the \bibtex{} to EndNote example, we encountered lenses of the form $\project{R}{s}$,

\noindent $\squash{c}{c'}{f}$ and $\perm{c_1, \ldots, c_n}{c}$ QREs, as well as
the regular operators applied to these, as in\bcp{bad linebreak: rewrite}

@@

One combinator that we haven't yet introduced\bcp{haven't we? we've
  mentioned it, at least.} is sequential composition.
The $c \semicolon c'$ QRE successively applies
two equivalence relations to a regular language. For example, suppose that
we wish to express a name as $\kw{last\_name \cdot ``,\text{''} \cdot
\text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot last\_name}$ as the
canonical representation. We also wish to project away the extra whitespace in
between \kw{first\_name} and \kw{last\_name} once this identification has been
made. Then we can express this equivalence relation by a composition of a
\kw{squash} QRE and a QRE that projects away the extra whitespace in between
\kw{first\_name} and \kw{last\_name}: \bcp{It would be easier to distinguish
code examples from text if they were indented a bit.}

@@

The final combinator is $\normalize{R}{R'}{f}$, which expresses
the equivalence relation that sends each element in $\mathcal{L}(R)$ to its
canonical representation $f(r) \in \mathcal{L}(R')$. The equivalence relation
defined by $\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the ``{fibres}'' of $f$---i.e., $r \sim r'$ if and
only if $f(r) = f(r')$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that, for any QRE
$q$, there exist regular expressions $R$ and $R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations that are either
difficult to define, or that cannot be defined using the other combinators.
\saz{Add something about being ``complete'' here?}\sam{I'm not sure how to do
this here. The precise statement about how normalize is strong enough to
express all QREs requires that we introduce W(c), K(c) and Canonizer(c) which
we haven't introduced at this point in the paper}\bcp{But we can at least
state an informal claim and give a forward pointer to where it is made more
precise later.  Or, since these definitions appear quite soon, perhaps we
can just reorganize the material a bit so that the definitions come before
the intuition plus a precise claim.}

Now in order for the programmer to ensure that the {\em normalize} QRE is well
formed, she must define a canonizing function $f$ between two regular
languages and prove that $f$ is surjective and idempotent: the compiler cannot
verify this since this problem is undecidable in general. Consequently, we do
not use the $\normalize{R}{R'}{f}$ QRE in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well behaved,'' particularly with respect to the strategy for
synthesizing quotient lenses that we will introduce in Section~\ref{synth}.

The equivalence relations definable by QREs are a strict subset of those that
can be expressed in Boomerang. For instance, in Boomerang, one can use any lens
to canonize a regular language described by $R$ to a regular language described
by $S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE
$c$ canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$. QREs therefore are
not as expressive as Boomerang canonizers.\bcp{Not sure I understand what
  this means, practically.  First, is it true that QREs can express any
  boomerang canonizer that maps from a language to a subset of the same
  language?  And second, can't at least some Boomerang canonizers be expressed as the
  composition of a QRE canonizer from R to R' plus a lens from R' to S?} On
the other hand, QREs enable us to 
express many equivalences that occur in real world data as we shall demonstrate
in the evaluation section (Section~\ref{impl}).

The net effect of these design decisions means that each QRE $c$ enables
us to simultaneously express
\begin{enumerate}
  \item a regular expression $W(c)$ (the ``whole'' of $c$),
  \item an equivalence relation $\eqrel{c}$ on $\mathcal{L}(W(c))$,
  \item a regular expression $K(c)$ (the ``kernel'' of $c$)
  such that $\mathcal{L}(K(c))$ forms a complete set of representatives for
  $\eqrel{c}$, and
  \item a ``canonizing'' function $\canonizer(c):\mathcal{L}(W(c))
  \longrightarrow \mathcal{L}(K(c))$ which given any $w \in \mathcal{L}(W(c))$,
  computes $\canonizer(c)(w)$ as the unique $k$ in $\mathcal{L}(K(c))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{c}$
\end{enumerate}
Observe that since $\canonizer(c)$ computes $\canonizer(c)(w)$ as the unique
$k$ in $\mathcal{L}(K(c))$ such that $w$ is equivalent to $k$ mod $\eqrel{c}$,
then the equivalence classes of $\eqrel{c}$ are the same as the
fibres\bcp{this is not a very familiar term for lots of people; if we're
  going to use it a lot, we should define it more carefully} of
$\canonizer(c)$.
